Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-07-03 _tex/paper2.tex
Built site for gh-pages

\begin{abstract}
This study revisits Thurstone's law of comparative judgment (CJ),
focusing on two prominent issues of traditional approaches. First, it
critiques the heavy reliance on Thurstone's Case V assumptions and, by
extension, the Bradley-Terry-Luce (BTL) model when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify the trait measurement model, they may fail to
capture the complexity of CJ data, potentially leading to unreliable and
inaccurate trait estimates. Second, the study highlights the apparent
disconnect between CJ's trait measurement and hypothesis testing
processes. Although separating these processes simplifies the analysis
of CJ data, it may also undermine the reliability of various statistical
results derived from these processes.

To address these issues, the study extends Thurstone's general form
using a systematic and integrated approach based on Causal and Bayesian
inference methods. This extension integrates core theoretical principles
alongside key CJ assessment design features, such as the selection of
judges, stimuli, and comparisons. It then translates these elements into
a probabilistic statistical model for analyzing dichotomous CJ data,
overcoming the rigid assumptions of Case V and the BTL model.

@@

In \emph{comparative judgment} (CJ) studies, judges assess a specific
trait or attribute across different stimuli by performing pairwise
comparisons \citep{Thurstone_1927a, Thurstone_1927b}. Each comparison
produces a dichotomous outcome, indicating which stimulus is perceived
to have a higher trait level. For example, when assessing writing
quality, judges compare pairs of written texts (the stimuli) to
determine the relative writing quality each text exhibit (the trait)
\citep{Pollitt_2012b, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

Numerous studies have documented the effectiveness of CJ in assessing
traits and competencies over the past decade. These studies have
highlighted three aspects of the method's effectiveness: its
reliability, validity, and practical applicability. Research on
reliability suggests that CJ requires a relatively modest number of
pairwise comparisons \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}
to generate trait scores that are as precise and consistent as those
generated by other assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}. In
addition, the evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
Meanwhile, research on validity indicates the capacity of CJ scores to
represent accurately the traits under measurement
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
Lastly, research on its practical applicability highlights CJ's
versatility across both educational and non-educational contexts
\citep{Kimbell_2012, Jones_et_al_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020b, Boonen_et_al_2020}.

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues of traditional approaches that can undermine the
reliability and validity of CJ's trait estimates. First, it critiques
the heavy reliance on Thurstone's Case V assumptions
\citep{Thurstone_1927b} and, by extension, the Bradley-Terry-Luce (BTL)
model \citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify the trait measurement model, they may fail to
capture the complexity of some traits or account for heterogeneous
stimuli, potentially leading to unreliable and inaccurate trait
estimates. Second, the study highlights the disconnect between CJ's
trait measurement and hypothesis testing processes. Although separating
these processes simplifies the analysis of CJ data, it may also
undermine the reliability of various statistical inferences derived from
these processes
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.

To address these issues, this study extends Thurstone's general form
through a systematic and integrated approach that combines causal and
Bayesian inference methods. In addition to potentially enhancing
measurement reliability and validity, and improving statistical accuracy
in hypothesis testing, this approach offers two key advantages. First,
it clarifies the interactions among all actors and processes involved in
CJ assessments. Second, it shifts the current comparative data analysis
paradigm from passively accepting Case V and the BTL model assumptions
to actively testing whether those assumptions fit the data under
analysis.

The study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} extends Thurstone's general
form to address these challenges. The extension integrates core
theoretical principles alongside key CJ assessment design features, such
as the selection of judges, stimuli, and comparisons.
Section~\ref{sec-statistical} translates these theoretical and practical
elements into a probabilistic statistical model to analyze dichotomous
pairwise comparison data. Finally, Section~\ref{sec-discussion}
discusses the findings, explores avenues for future research, and detail
the challenges for future researchers.

@@

In its most general form, Thurstone's theory addresses pairwise
comparisons wherein a single judge evaluates multiple stimuli
\citep{Thurstone_1927b}. The theory posits that two key factors
determine the dichotomous outcome of these comparisons: the discriminal
process of each stimulus and their discriminal difference. The
\emph{discriminal process} captures the psychological impact each
stimulus exerts on the judge or, more simply, his perception of the
stimulus trait. The theory assumes that the discriminal process for any
given stimulus forms a Normal distribution along the trait continuum
\citep{Thurstone_1927b}. The mode (mean) of this distribution, known as
the \emph{modal discriminal process}, indicates the stimulus position on
this continuum, while its dispersion, referred to as the
\emph{discriminal dispersion}, reflects variability in the perceived
trait of the stimulus.

Figure~\ref{fig-discriminal_process} illustrates the hypothetical
discriminal processes along a quality trait continuum for two written
texts. The figure indicates that the modal discriminal process for Text
B is positioned further along the continuum than that of Text A
\((T_{B} > T_{A})\), suggesting that Text B exhibits higher quality.
Additionally, the figure highlights that Text B has a broader
distribution compared to Text A, which arises from its larger
discriminal dispersion \((\sigma_{B} > \sigma_{A})\).

@@

\subcaption{\label{fig-discriminal_process}Discriminal processes}

@@

\subcaption{\label{fig-discriminal_difference}Discriminal difference}

@@

However, since the individual discriminal processes of the stimuli are
not directly observable, the theory introduces the \emph{law of
comparative judgment}. This law posits that in pairwise comparisons, a
judge perceives the stimulus with a discriminal process positioned
further along the trait continuum as possessing more of the trait
\citep{Bramley_2008}. This suggests that pairwise comparison outcomes
depend on the relative distance between stimuli, not their absolute
positions on the continuum. Indeed, the theory assumes that the
difference between the underlying discriminal processes of the stimuli,
referred to as the \emph{discriminal difference}, determines the
observed dichotomous outcome. Furthermore, the theory assumes that
because the individual discriminal processes form a Normal distribution
on the continuum, the discriminal difference will also conform to a
Normal distribution \citep{Andrich_1978}. In this distribution, the mode
(mean) represents the average relative separation between the stimuli,
and its dispersion indicates the variability of that separation.

@@

Thurstone noted from the outset that his general formulation, described
in Section~\ref{sec-thurstone_theory}, led to a \emph{trait scaling
problem}. Specifically, the model required estimating more ``unknown''
parameters than the number of available pairwise comparisons
\citep{Thurstone_1927b}. For instance, in a CJ assessment with five
texts, the general form would require estimating \(20\) parameters: five
modal discriminal processes, five discriminal dispersions, and \(10\)
correlations--one per comparison (see Table~\ref{tbl-thurstone_cases}).
However, a single judge could only provide \({5 \choose 2} = 10\) unique
comparisons, an insufficient data set to estimate the required
parameters.

@@

To address this issue and facilitate the practical implementation of the
theory, Thurstone developed five cases derived from this general form,
each progressively incorporating additional simplifying assumptions
\citep{Thurstone_1927b}. In Case I, Thurstone postulated that pairs of
stimuli would maintain a constant correlation across all comparisons. In
Case II, he allowed multiple judges to undertake comparisons instead of
confining evaluations to a single judge. In Case III, he posited that
there was no correlation between stimuli. In Case IV, he assumed that
the stimuli exhibited similar dispersions. Finally, in Case V, he
replaced this assumption with the condition that stimuli had equal
discriminal dispersions. Table~\ref{tbl-thurstone_cases} summarizes the
assumptions of the general form and the five cases. For a detailed
discussion of these cases and their progression, refer to
\citet{Thurstone_1927b} and \citet{Bramley_2008}.

However, Thurstone developed Case V prioritizing statistical simplicity
over precise trait measurement and offering no guidance on how to use
its trait estimates for statistical inference or hypothesis testing.
Specifically, Thurstone cautioned that its use ``should not be made
without (an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, as it
imposes the most extensive set of simplifying assumptions
\citep{Bramley_2008, Kelly_et_al_2022} (see
Table~\ref{tbl-thurstone_cases}). Moreover, because Thurstone's primary
goal was to produce a ``rather coarse scaling'' of traits and ``allocate
the compared stimuli on this continuum''
\citep[pp.~269]{Thurstone_1927b}, his theory did not support formal
statistical inference. Despite these limitations, CJ research has
predominantly relied on Case V to measure different traits, which raises
significant concerns about the reliability and validity of such
measurements in contexts where the case's assumptions may not hold
\citep{Kelly_et_al_2022, Andrich_1978}. Furthermore, although the CJ
tradition has attempted to address the gap in hypothesis testing by
relying on the point estimates of traits--or their transformations--the
statistical literature cautions against using these estimates as the
sole basis for statistical inference, as such practices introduce bias
in the analysis and reduce the precision of hypothesis tests
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}. Next, both
issues are discussed in more depth.

@@

As previously discussed, Case V remains the most widely used model in CJ
literature. This preference largely stems from the widespread adoption
of the BTL model, which provides a simplified statistical representation
of the case. The BTL model mirrors most of Case V's assumptions, with
one notable distinction. While Case V assumes a Normal distribution for
the stimuli' discriminal processes, the BTL model uses the more
mathematically tractable Logistic distribution
\citep{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). However, this substitution has minimal
impact on trait estimation or model interpretation because the scale of
the discriminal process (i.e., the latent trait) is arbitrary up to a
non-monotonic transformation
\citep{vanderLinden_et_al_2017_I, McElreath_2021}. That is, as long as
the substitution (transformation) preserves the data rank order, the
choice of distribution for the discriminal processes is inconsequential.
This condition is satisfied in this case, as the Normal and Logistic
distributions exhibit analogous statistical properties, differing only
by a scaling factor of approximately \(1.7\)
\citep{vanderLinden_et_al_2017_I}.

However, Thurstone acknowledged that some assumptions of Case V could be
problematic when researchers assess complex traits or heterogeneous
stimuli \citep{Thurstone_1927a}. Thus, given that modern CJ applications
often involve such traits and stimuli, two key assumptions of Case V,
and by extension, the BTL model, may not always hold in theory or
practice. These assumptions are the equal dispersion and zero
correlation between stimuli.

@@

According to the theory, discrepancies in the discriminal dispersions of
stimuli shape the distribution of the discriminal difference, directly
influencing the outcome of pairwise comparisons. A thought experiment
can help illustrate this idea. In it, researchers observe the
discriminal processes for two texts, A and B, assuming that the
dispersion for Text A remains constant and that the two texts are
uncorrelated \((\rho=0)\). Figure~\ref{fig-dispersion} demonstrates that
an increase in the uncertainty associated with the perception of Text B
relative to Text A \((\sigma_{B} - \sigma_{A})\), broadens the
distribution of their discriminal difference. This broadening affects
the probability area where the discriminal difference distinctly favors
Text B over Text A, expressed as \(P(B>A)\), ultimately influencing the
comparison outcome. Additionally, the figure reveals that when the
discriminal dispersions of the texts are equal, as in the BTL model
\((\sigma_{B} - \sigma_{A}=0)\), the discriminal difference distribution
is more narrow than when the dispersions differ. As a result, the
discriminal difference is more likely to favor Text B over Text A, as it
is represented by the shaded gray area.

@@

In experimental practice, however, the thought experiment occurs in
reverse. Researchers first observe the comparison outcome and then use
the BTL model to infer the discriminal difference between stimuli and
their respective discriminal processes \citep{Thurstone_1927a}.
Consequently, the outcome's ability to reflect \emph{true} differences
between stimuli largely depends on the validity of the model's
assumptions \citep{Kohler_et_al_2019}, in this case, the assumption of
equal dispersions. When the assumption accurately captures the
complexity of the data, the BTL model estimates a discriminal difference
distribution that accurately represents the \emph{true} discriminal
difference between the texts. This scenario is illustrated in
Figure~\ref{fig-dispersion}, when the model's discriminal difference
distribution aligns with the \emph{true} discriminal difference
distribution, represented by the thick continuous line corresponding to
\(\sigma_{B}-\sigma_{A}=0\). The accuracy of this discriminal difference
then ensures reliable estimates for the texts' discriminal processes.

Notably, while assuming equal dispersions simplifies the trait
measurement model, evidence from the CJ literature suggests that this
assumption may fail to accommodate for heterogeneous stimuli, such as
handwritten texts or English compositions
\citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022}.
The presence of the so-called \emph{misfit texts}--texts that elicit
more judgment discrepancies than others
\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018}--may
signal this limitation, as such discrepancies can arise from larger
discriminal dispersions due to stimulus heterogeneity or because the
texts are genuine outliers--that is, texts with distinctive
characteristics that deviate markedly from the rest of the sample
\citep{Grubbs_1969}. In either case, the BTL model's assumptions prevent
it from adequately accounting for or addressing these anomalies, leaving
exclusion of the ``problematic'' texts as the primary remedy
\citep{Pollitt_2012a, Pollitt_2012b}.

@@

Despite these notable differences in the distribution of the discriminal
difference under various correlational assumptions, in practice,
assessment designs often adopt the assumption of no correlation between
stimuli based on an old theoretical justification. Specifically,
\citet{Thurstone_1927b} argued that stimuli could be treated as
uncorrelated because judges' biases--arising from two opposing and
equally weighted effects occurring during the pairwise
comparisons--would cancel each other out. This idea was later formalized
by \citet{Andrich_1978}, who provided a mathematical demonstration of
this cancellation using the BTL model under the assumption of
discriminal processes with additive biases. However, evidence from the
CJ literature indicates that the assumption of zero correlation does not
hold in practice in at least two cases: when intricate aspects of
multidimensional, complex traits or heterogeneous stimuli influence
judges' perceptions or when additional hierarchical structures are
relevant to the stimuli.

Research on text quality assessments suggests that when judges evaluate
complex, multidimensional traits or heterogeneous stimuli, they often
rely on a variety of intricate stimulus characteristics to inform their
judgments
\citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022}.
Regardless of their relevance, these characteristics may not receive
equal weight or consistently oppose one another across comparisons. As a
result, they may exert a disproportionate influence on judges'
perceptions, generating biases that persist rather than cancel out. For
example, this could occur when a judge assessing the argumentative
quality of a text may place disproportionate emphasis on handwriting
clarity, thereby favoring neatly written texts despite their weaker
arguments. Moreover, because the discriminal process of stimuli becomes
an observable outcome only through the judges' perceptions, these biases
could introduce dependencies between the stimuli
\citep{vanderLinden_et_al_2017_II}. While direct evidence for this exact
scenario is limited, existing studies document the presence of judge
bias in CJ
\citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a},
reinforcing the argument that the factors influencing pairwise
comparisons do not always cancel each other out.

In the second case, the shared context or inherent connections
introduced by additional hierarchical structures may create dependencies
between stimuli--a statistical phenomenon known as clustering
\citep{Everitt_et_al_2010}. For instance, when the same individual
produces multiple texts, those texts often share several features such
as writing style or overall quality. Although the CJ literature
acknowledges the existence of such hierarchical structures
\citep[e.g.,][]{Boonen_et_al_2020}, the statistical approaches to
account for this additional source of dependence have been insufficient.
For instance, when CJ data incorporates multiple samples of stimuli from
the same individuals, researchers frequently rely on (averaged) point
estimates of the BTL scores to conduct subsequent analyses and tests at
the individual level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

>>>>

\begin{abstract}
This study revisits Thurstone's law of comparative judgment (CJ) by
addressing two central issues in the CJ literature. First, it critiques
the widespread reliance on Thurstone's Case V assumptions and, by
extension, the Bradley-Terry-Luce (BTL) model in the analysis of CJ
data. Specifically, this study argues that while the assumptions of
equal discriminal dispersions and zero correlation between stimuli
simplify the trait measurement model, they may fail to capture the
complexity of some traits or account for heterogeneous stimuli,
potentially leading to unreliable and inaccurate trait estimates.
Second, the study highlights the apparent disconnect between trait
measurement and hypothesis testing procedures in CJ applications. It
contends that while separating these procedures simplifies the analysis
of CJ data, this practice can also undermine the resulting statistical
inferences.

To address these issues, this study aims to extend Thurstone's general
form using a systematic and integrated approach based on causal and
Bayesian inference methods. This extension integrates Thurstone's core
theoretical principles alongside key CJ assessment design features. It
then translates these elements into a probabilistic statistical model
for analyzing dichotomous CJ data, overcoming the rigid assumptions of
Case V and the BTL model.

@@

In \emph{comparative judgment} (CJ) studies, judges assess a specific
trait or attribute across different stimuli by performing pairwise
comparisons \citep{Thurstone_1927a, Thurstone_1927b}. Each comparison
produces a dichotomous outcome, indicating which stimulus is perceived
to have a higher trait level. For example, judges can compare pairs of
written texts (the stimuli) to determine the relative writing quality
each text exhibit (the trait)
\citep{Pollitt_2012b, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

Numerous studies have documented the effectiveness of CJ in assessing
traits and competencies over the past decade. These studies have
highlighted three key strengths of the method: its reliability, its
validity, and practical applicability. Research on reliability suggests
that CJ requires a relatively modest number of pairwise comparisons
\citep{Verhavert_et_al_2019, Crompvoets_et_al_2022} to generate trait
scores that are as precise and consistent as those generated by other
assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}. In
addition, the evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
Meanwhile, research on validity indicates the capacity of CJ scores to
represent accurately the traits under measurement
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
Lastly, research on its practical applicability highlights CJ's
versatility across educational and non-educational contexts
\citep{Kimbell_2012, Jones_et_al_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020b, Boonen_et_al_2020}.

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues in the CJ literature that can undermine the reliability
and validity of CJ's trait estimates. First, it critiques the widespread
reliance on Thurstone's Case V assumptions \citep{Thurstone_1927b} and,
by extension, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} in the analysis of CJ data.
Specifically, this study argues that while the assumptions of equal
discriminal dispersions and zero correlation between stimuli simplify
the trait measurement model, they may fail to capture the complexity of
some traits or account for heterogeneous stimuli, potentially leading to
unreliable and inaccurate trait estimates. Second, the study highlights
the disconnect between trait measurement and hypothesis testing
procedures in CJ applications. In particular, it contends that while
separating these procedures simplifies the analysis of CJ data, this
practice can also undermine the resulting statistical inferences
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.

To address these issues, this study aims to extend Thurstone's general
form through a systematic and integrated approach that combines causal
and Bayesian inference methods. This extension integrates Thurstone's
core theoretical principles alongside key CJ assessment design features,
such as the selection of judges, stimuli, and comparisons. In addition
to potentially enhancing measurement reliability and validity, and
improving statistical accuracy in hypothesis testing, this approach
offers two key advantages. First, it clarifies the interactions among
all actors and processes involved in CJ assessments. Second, it shifts
the current comparative data analysis paradigm from passively accepting
Case V and the BTL model assumptions to actively testing whether those
assumptions fit the data under analysis.

The remainder of this study is organized into seven sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} examines the central issues
identified in the CJ literature. Section~\ref{sec-theoretical} extends
Thurstone's general form to address these issues. The extension
integrates core theoretical principles alongside key CJ assessment
design features, such as the selection of judges, stimuli, and
comparisons. Section~\ref{sec-statistical} translates these theoretical
and design elements into a probabilistic statistical model to analyze
dichotomous pairwise comparison data. Section~\ref{sec-discussion}
reviews the findings, outline directions for future research, discusses
the study's limitations and details the challenges that applied
researchers may encounter. Finally, Section~\ref{sec-conclusion}
provides the concluding remarks.

@@

In its \emph{general form}, Thurstone's theory addresses pairwise
comparisons of a single judge who evaluates multiple stimuli
\citep{Thurstone_1927a, Thurstone_1927b}. The theory posits that two key
factors determine the dichotomous outcome of these comparisons: the
discriminal process of each stimulus and their discriminal difference.
The \emph{discriminal process} captures the psychological impact each
stimulus exerts on the judge or, more simply, his perception of the
stimulus trait. The theory assumes that the discriminal process for any
given stimulus forms a Normal distribution along the trait continuum
\citep{Thurstone_1927b}. The mode (mean) of this distribution, known as
the \emph{modal discriminal process}, indicates the stimulus position on
this continuum, while its dispersion, referred to as the
\emph{discriminal dispersion}, reflects variability in the perceived
trait of the stimulus.

These ideas become clearer with an example.
Figure~\ref{fig-discriminal_process} illustrates the hypothetical
discriminal processes for two written texts along a \emph{quality} trait
continuum. The figure shows that the modal discriminal process for Text
B lies further along the continuum than that of Text A
\((T_{B} > T_{A})\), suggesting that Text B exhibits higher quality. The
figure also shows that Text B has a broader distribution than Text A,
indicating a greater variability in perception due to its larger
discriminal dispersion \((\sigma_{B} > \sigma_{A})\).

@@

\subcaption{\label{fig-discriminal_process}Hypothetical discriminal
processes for two written texts}

@@

\subcaption{\label{fig-discriminal_difference}Hypothetical discriminal
difference of two written texts}

@@

However, since the individual discriminal processes of the stimuli are
not directly observable \citep{Thurstone_1927a}, the theory introduces
the \emph{law of comparative judgment}. This law posits that in pairwise
comparisons, a judge perceives the stimulus with a discriminal process
positioned further along the trait continuum as possessing more of the
trait \citep{Bramley_2008}. This suggests that pairwise comparison
outcomes depend on the relative distance between stimuli, not their
absolute positions on the continuum. Indeed, the theory assumes that the
difference between the underlying discriminal processes of the stimuli,
referred to as the \emph{discriminal difference}, determines the
observed dichotomous outcome. Furthermore, the theory assumes that
because the individual discriminal processes form a Normal distribution
on the continuum, the discriminal difference will also conform to a
Normal distribution \citep{Andrich_1978}. In this distribution, the mode
(mean) represents the average relative separation between the stimuli
\((T_{B} - T_{A})\), and its dispersion \(\sigma_{BA}\) indicates the
variability of that separation.

@@

Thurstone noted from the outset that his general formulation, described
in Section~\ref{sec-thurstone_theory}, led to a \emph{trait scaling
problem}. Specifically, the model required estimating more ``unknown''
parameters than the number of available pairwise comparisons
\citep{Thurstone_1927b}. For instance, in a CJ assessment with five
texts, the general form would require estimating \(20\) parameters: five
modal discriminal processes, five discriminal dispersions, and \(10\)
correlations--one per comparison. However, a single judge could only
provide \({5 \choose 2} = 10\) unique comparisons, an ``insufficient''
data set to estimate the required parameters.

To address this issue and facilitate the practical implementation of the
theory, Thurstone developed five cases derived from this general form,
each progressively incorporating additional simplifying assumptions
\citep{Thurstone_1927b}. In Case I, Thurstone postulated that pairs of
stimuli would maintain a constant correlation across all comparisons. In
Case II, he allowed multiple judges to undertake comparisons instead of
confining evaluations to a single judge. In Case III, he posited that
there was no correlation between stimuli. In Case IV, he assumed that
the stimuli exhibited similar dispersions. Finally, in Case V, he
replaced this assumption with the condition that stimuli had equal
discriminal dispersions. Table~\ref{tbl-thurstone_cases} summarizes the
assumptions of the general form and the five cases. For a detailed
discussion of these cases and their growing simplification, refer to
\citet{Thurstone_1927b} and \citet{Bramley_2008}.

@@

As the table suggest, Thurstone developed Case V with an emphasis on
statistical simplicity, but this simplicity comes at the expense of
accurate and precise trait measurement and practical guidance for
inference. Specifically, Thurstone cautioned that Case V use ``should
not be made without (an) experimental test''
\citep[p.~270]{Thurstone_1927b}, as it imposes the most extensive set of
simplifying assumptions \citep{Bramley_2008, Kelly_et_al_2022} (see
Table~\ref{tbl-thurstone_cases}). Moreover, because Thurstone's primary
goal was to produce a ``rather coarse scaling'' of traits and ``allocate
the compared stimuli on this continuum''
\citep[p.~269]{Thurstone_1927b}, his theory did not support formal
statistical inference. Nevertheless, despite Case V's limitations, CJ
research has predominantly relied on it to measure various traits,
raising significant concerns about the reliability and validity of such
measurements in contexts where its assumptions may not hold
\citep{Kelly_et_al_2022, Andrich_1978}. Furthermore, although the CJ
tradition has attempted to address the gap in statistical inference by
relying on the point estimates of traits--or their transformations--the
statistical literature cautions against using these estimates as the
sole basis for inference, as such practices introduce bias and reduce
the precision of hypothesis tests
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}. The next
sections discuss both issues in greater depth.

@@

Case V is the most widely used model in the CJ literature. This
preference largely stems from the widespread adoption of the BTL model,
which provides a simplified statistical representation of the case. The
BTL model incorporates most of Case V's assumptions, with one notable
exception. While Case V assumes a Normal distribution for the stimuli'
discriminal processes, the BTL model uses the more mathematically
tractable Logistic distribution \citep{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). However, this substitution has minimal
impact on trait estimation or model interpretation because the scale of
the discriminal process (i.e., the latent trait) is arbitrary up to a
non-monotonic transformation
\citep{vanderLinden_et_al_2017_I, McElreath_2021}. In other words, as
long as the substitution (transformation) preserves the rank order of
the data, the choice of distribution for the discriminal processes is
inconsequential. The BTL model satisfies this condition because the
Normal and Logistic distributions exhibit analogous statistical
properties, differing only by a scaling factor of approximately \(1.7\)
\citep{vanderLinden_et_al_2017_I}.

However, Thurstone acknowledged that some assumptions of Case V are
problematic when researchers assess complex traits or heterogeneous
stimuli \citep{Thurstone_1927a}. Thus, given that contemporary CJ
applications often involve such traits and stimuli, two key assumptions
of Case V, and by extension, the BTL model, may not always hold in
theory or practice. These assumptions are the equal dispersion and zero
correlation between stimuli.

@@

According to the theory, discrepancies in the discriminal dispersions of
stimuli shape the distribution of the discriminal difference, directly
influencing the outcome of pairwise comparisons. A thought experiment
can help illustrate this idea. Suppose researchers observe the
discriminal processes for two texts, A and B, assuming that the
dispersion for Text A remains constant and that the two texts are
uncorrelated \((\rho=0)\). Figure~\ref{fig-dispersion} demonstrates that
an increase in the uncertainty associated with the perception of Text B
relative to Text A \((\sigma_{B} - \sigma_{A})\), broadens the
distribution of their discriminal difference. This broadening affects
the probability area where the discriminal difference distinctly favors
Text B over Text A, expressed as \(P(B>A)\), ultimately influencing the
comparison outcome. Additionally, the figure reveals that when the
discriminal dispersions of the texts are equal, as in the BTL model
\((\sigma_{B} - \sigma_{A}=0)\), the discriminal difference distribution
is more narrow than when the dispersions differ. As a result, the
discriminal difference is more likely to favor Text B over Text A, as it
is represented by the shaded gray area.

@@

In experimental practice, however, the thought experiment occurs in
reverse. Researchers first observe the comparison outcome and then use
the BTL model to infer the discriminal difference between stimuli and
their respective discriminal processes \citep{Thurstone_1927a}.
Consequently, the effectiveness of the outcome to reflect \emph{true}
differences between stimuli largely depends on the validity of the
model's assumptions \citep{Kohler_et_al_2019}, in this case, the
assumption of equal dispersions. When the assumption accurately captures
the complexity of the data, the BTL model estimates a discriminal
difference distribution that accurately represents the \emph{true}
discriminal difference between the texts. This scenario is illustrated
in Figure~\ref{fig-dispersion}, when the model's discriminal difference
distribution aligns with the \emph{true} discriminal difference
distribution, represented by the thick continuous line corresponding to
\(\sigma_{B}-\sigma_{A}=0\). The accuracy of this discriminal difference
then ensures reliable estimates for the texts' discriminal processes.

Notably, while assuming equal dispersions simplifies the trait
measurement model, evidence from the CJ literature suggests that this
assumption may not hold for heterogeneous stimuli, such as handwritten
texts or English compositions
\citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022}.
The presence of the so-called misfit texts signals this limitation.
\emph{Misfit texts} are those that elicit more judgment discrepancies
than others
\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018},
and these discrepancies may arise from larger discriminal dispersions
caused by the stimulus' heterogeneity or because the texts are genuine
outliers--i.e., texts with distinctive characteristics that deviate
markedly from the rest of the sample in which they occur
\citep{Grubbs_1969}. In either case, the BTL model's assumptions prevent
it from adequately accounting for or addressing these anomalies, leaving
exclusion of such ``problematic'' texts as the primary remedy
\citep{Pollitt_2012a, Pollitt_2012b}.

@@

Despite these notable differences in the distribution of the discriminal
difference under various correlational assumptions, in practice,
assessment designs often adopt the assumption of no correlation between
stimuli based on Thurstone's early theoretical justification
\citep{Thurstone_1927b}. He argued that stimuli could be treated as
uncorrelated because judges' biases--arising from two opposing and
equally weighted effects occurring during the pairwise
comparisons--would cancel each other out. This idea was later formalized
by \citet{Andrich_1978}, who provided a mathematical demonstration of
this cancellation using the BTL model under the assumption of
discriminal processes with additive biases. However, evidence from the
CJ literature indicates that the assumption of zero correlation does not
hold in practice in at least two cases: (1) when intricate aspects of
multidimensional, complex traits or heterogeneous stimuli influence
judges' perceptions, (2) when additional hierarchical structures are
relevant to the stimuli.

Regarding the first case, research on text quality assessments suggests
that when judges evaluate complex, multidimensional traits or
heterogeneous stimuli, they often rely on a variety of intricate
stimulus characteristics to inform their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022}.
Regardless of their relevance, these characteristics may not be equally
weighted or consistently opposed across comparisons. As a result, they
may exert a disproportionate influence on judges' perceptions,
generating biases that persist rather than cancel out. For example, this
might occur when a judge assessing the argumentative quality of a text
is disproportionately influenced by the clarity of the handwriting,
thereby favoring neatly written texts even if their arguments are
weaker. Moreover, because the discriminal process of stimuli becomes an
observable outcome only through the judges' perceptions, these biases
could introduce dependencies between the stimuli
\citep{vanderLinden_et_al_2017_II}. While direct evidence for this exact
scenario is limited, existing studies document the presence of judge's
biases in CJ contexts
\citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a},
reinforcing the argument that the factors influencing pairwise
comparisons do not always cancel each other out.

In the second case, the shared context or inherent connections
introduced by additional hierarchical structures may create dependencies
between stimuli--a statistical phenomenon known as clustering
\citep{Everitt_et_al_2010}. For instance, when the same individual
produces multiple texts, those texts often share several features such
as writing style or overall quality. Although some CJ studies
acknowledge the presence of such hierarchical structures and account for
them \citep[e.g.,][]{Boonen_et_al_2020}, the treatment of this
additional source of dependence in other research has often been
insufficient. For instance, when CJ data include multiple samples of
stimuli from the same individuals, researchers frequently rely on
(averaged) point estimates of the BTL scores to conduct subsequent
analyses and tests at the individual level
\citep{Bramley_et_al_2019, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

Thus, erroneously assuming zero correlation between stimuli can also
lead to significant statistical and measurement issues. In particular,
neglecting judges' biases or relevant hierarchical structures can create
dimensional mismatches in the model, leading to the over- or
underestimation of trait reliability
\citep{Ackerman_1989, Hoyle_et_al_2023} and even introduce statistical
biases \citep{Wu_et_al_2022}. These inaccuracies can result in spurious
conclusions about the discriminal differences \citep{McElreath_2020}
and, by extension, the underlying discriminal processes of the stimuli.
One such spurious conclusion could be the incorrect classification of
stimuli (or judges) as \emph{misfits}. Figure~\ref{fig-correlation}
illustrates how assuming zero correlation can undermine trait
reliability: the discriminal difference distribution of the BTL scores
follows the thick continuous line \((\rho = 0)\), while the \emph{true}
discriminal difference may correspond to any discontinuous line where
\(\rho \neq 0\).
